Phylogenetics

### 2024-03-27 episodes/phylogenetics.Rmd
Progressed with blast episod.
Switched the phylogenetics episode to Uppmax and phylo.io

## Setup

The exercise is done for a large part with pen and paper, and then a 
demonstration in R. We'll also use the R package 
[`ape`](https://emmanuelparadis.github.io/) , which you should
install if it's not present on your setup:

@@

## UPGMA-based tree

@@

## Calculating distance "by hand"

@@

## Introduction

@@

## Neighbor-joining

@@

To perform NJ on our sequences, we'll use the function inbuilt in Seaview. First (re-)open the alignment we obtained from `mafft` with the E-INS-i method.

@@

```bash
seaview rpoB.einsi.aln &
```

@@

## Maximum likelihood

We will now use IQ-TREE to infer a maximum-likelihood tree of the RpoB dataset we aligned with `mafft` previously.

```bash
$ iqtree -s rpoB.einsi.aln -m TEST -B 1000
```

Here, we tell IQ-TREE to use the alignment `rpoB.einsi.aln`, and to test among the standard substitution models which one fits best (`-m TEST`). We also tell IQ-TREE to perform 1000 ultra-fast bootstraps (`-B 1000`). We'll discuss these later.

@@

```bash
ls rpoB.einsi.aln.*
```

```output
rpoB.einsi.aln.bionj      rpoB.einsi.aln.iqtree     rpoB.einsi.aln.model.gz
rpoB.einsi.aln.ckp.gz     rpoB.einsi.aln.log        rpoB.einsi.aln.splits.nex
rpoB.einsi.aln.contree    rpoB.einsi.aln.mldist     rpoB.einsi.aln.treefile
```

@@

Now display the resulting tree in FigTree.

```bash
$ figtree rpoB.einsi.aln.treefile &
```

A window will pop-up and ask to provide a name for branch/node labels. Use e.g. 'bootstraps'.

The tree appears as unrooted. It is good practice to start by ordering the nodes and root it. Since we don't really know otherwise where the root is, we'll do a mid-point rooting to start with.

In the left menu, develop the 'Trees' menu. Tick the 'Root tree' and select 'Midpoint' from the drop-down menu. Tick the 'Order nodes' box and keep the 'increasing' option selected.

@@

```bash
$ iqtree -s rpoB.fftns.aln -m TEST -B 1000
$ figtree rpoB.fftns.aln.treefile &
```

@@

## Bootstraps

We have inferred four trees:
* Two based on the alignment generated from the E-INS-i algorithm, two from the FFT-NS.
* Two inferred with the BioNJ algorithm and two with the ML algorithm (IQ-Tree)

Along the way, we've generated bootstraps for all our trees. Now show them on all four trees.
* For the Seaview trees, tick the 'Br support' box
* For the trees shown in FigTree, develop the 'Branch labels' menu, tick the corresponding box and select 'bootstraps' from the 'Display' drop-down menu.

@@

## Genetic drift

>>>>

## Exercise 1: Paper-and-pen phylogenetic tree

### Setup

The exercise is done for a large part with pen and paper, and then a demonstration in R on your laptop. You can already RStudio. We'll also use the R package [`ape`](https://emmanuelparadis.github.io/) , which you should install if it's not present on your setup:

@@

### UPGMA-based tree

@@

### Calculating distance "by hand"

@@

## Exercise 2: maximum-likelihood tree

@@

### Neighbor-joining

@@

To perform NJ on our sequences, we'll use the function inbuilt in Seaview. First (re-)open the alignment we obtained from `mafft` with the E-INS-i method, trimmed with ClipKIT.

@@

```bash
seaview rpoB.einsi.clipkit.aln &
```

@@

### Maximum likelihood

We will now use IQ-TREE to infer a maximum-likelihood (ML) tree of the RpoB dataset we aligned with `mafft` previously. While you performed the first trees on your laptop, you will infer the ML tree on Uppmax. As usual, use the `interactive` command to gain access to compute time. Go to the `phylogentics` folder created in the MSA episode, and load the right modules.

:::::: challenge

Get to the right folder, require compute time and load the right modules.

::: solution

 ```bash
interactive -A uppmax2024-2-10 -M snowy -t 4:00:00
module load bioinfo-tools iqtree
cd /proj/g2020004/nobackup/3MK013/<username>/phylogenetics
```

::::::::::::

::::::::::::::::

Loading the `iqtree` module resulted in:

```output
iqtree/2.2.2.6-omp-mpi: executables are prefixed with 'iqtree2'. Additional modules are required for the multi-node MPI version. See 'module help iqtree/2.2.2.6-omp-mpi'.
```

You won't be using the multi-node MPI version, so you don't need to load additional modules. But you will need to use `iqtree2` instead of the "normal" `iqtree`.

Then run your first tree, on the `ClipKIT`-trimmed alignment.

```bash
$ iqtree2 -s rpoB.einsi.clipkit.aln -m TEST -B 1000
```

Here, we tell IQ-TREE to use the alignment `rpoB.einsi.clipkit.aln`, and to test among the standard substitution models which one fits best (`-m TEST`). We also tell IQ-TREE to perform 1000 ultra-fast bootstraps (`-B 1000`). We'll discuss these later.

@@

```bash
ls rpoB.einsi.clipkit.aln.*
```

```output
rpoB.einsi.clipkit.aln.bionj    rpoB.einsi.clipkit.aln.iqtree  rpoB.einsi.clipkit.aln.model.gz
rpoB.einsi.clipkit.aln.ckp.gz   rpoB.einsi.clipkit.aln.log     rpoB.einsi.clipkit.aln.splits.nex
rpoB.einsi.clipkit.aln.contree	rpoB.einsi.clipkit.aln.mldist  rpoB.einsi.clipkit.aln.treefile
```

@@

To visualize the tree, open the [Beta version of phylo.io](https://beta.phylo.io/viewer/#). Click on "add a tree" and copy/paste the content of the `treefile` (use e.g. `less` or `cat` to display it) in the box. Make sure the "Newick format" is selected and click on "Done".

The tree appears as unrooted. It is good practice to start by ordering the nodes and root it. There is no good way to automatically order the branches in phylo.io as of yet, but rerooting can be done by clicking on a branch and selecting "Reroot". Reroot the tree between archaea and bacteria. Now the tree makes a bit more sense.

@@

```bash
$ iqtree2 -s rpoB.fftns.aln -m TEST -B 1000
$ figtree rpoB.fftns.aln.treefile &
```

@@

### Bootstraps

We have inferred four trees:
* Two based on the alignment generated from the E-INS-i algorithm (trimmed by ClipKIT), two from the FFT-NS.
* Two inferred with the BioNJ algorithm and two with the ML algorithm (IQ-Tree)

Along the way, we've generated bootstraps for all our trees. Now show them on all four trees.
* For the Seaview trees, tick the 'Br support' box
* For the trees shown in phylo.io, click on "Settings" > "Branch & Labels" and above the branch, click the drop-down menu and select "Data".

@@

## Exercise 3: Genetic drift
